Chasing \rainbows and \oceanglint:\\ Inner working angle constraints for the Habitable Worlds Observatory

### 2023-05-11 src/tex/ms.tex
[showyourwork] overleaf sync

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\HWO), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight reflected by an exoplanet can reveal ocean glint, rainbows, and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, and \HWO's coronagraph may obscure the phase angles where such scattering features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
For planets in nearly edge-on orbits, accessible phase angles are limited by coronagraphic obscuration. 
%
We use \HWO'~catalogue of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering. 
%
We find that any exo-Earths in systems in this catalogue will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The rainbow due to water clouds at scattering angles of \qtyrange{140}{160}{\degree} would be accessible to \HWO\ in $\sim\num{42}$ systems, while ocean glint signature at scattering angles of $\qtyrange{50}{70}{\degree}$ would be accessible in $\sim\num{14}$ systems, assuming an \IWA of \qty{62}{\mas} ($3\,\lambda/D$ for a \qty{6}{\meter} telescope at \qty{600}{\nano\meter}).
%
Improving the \IWA to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases slightly for eccentric orbits.  %and scales approximately inversely with the \IWA.  
\end{abstract}

@@

\begin{figure*}
    \centering
    \includegraphics{figures/figure-1-bott-plot.pdf}
    \script{create-figure-1-bott-plot.py}
    \caption{
        The phase curves of the normalized polarized flux of a planet with a wavy ocean surface and an Earth-like atmosphere in an edge-on orbit (the planet thus exhibits phase angles ranging from \qty{0}{\degree} to \qty{180}{\degree} along its orbit). 
        The wind speed over the ocean is \qty{7}{\meter\per\second}. 
        The clouds in the planet's atmosphere consist of spherical water droplets. 
        The cloud coverage fraction is 0.5. 
        The cloud pattern is patchy, and the glint is kept cloud-free to maximize the glint signature \citep[see also Fig. 9 of][from which we took the mean signal of 300 planets with randomly generated patchy cloud patterns + 2 standard deviations, representative for cloud-free glint scenarios; the \enquote{noise} at the larger phase angles and longer wavelengths is due to the variation in the location of cloud patches]{treesandstam2019}. 
        The polarized flux is normalized to the total (unpolarized + polarized) flux (it is thus in fact the degree of polarization). 
%        Note that the reference plane for the polarized flux is chosen such that only the polarized components perpendicular or parallel to the plane differ significantly from zero.
 %   and the latter
 %   is equals the planet's geometric albedo at phase angle $\alpha = 0^\circ$.
        %The cloudy scenario includes a low-altitude, Earth-like water 
        %cloud deck. 
        Below the plot, we show the phase angle ranges that cover the
        characteristic features of glories, rainbows, Rayleigh scattering, glint, and forward scattering by cloud particles adopted in this paper. Note that the features in the actual curves partly overlap
        with each other.  
        %Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (phase angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest phase angles).
    }
    \label{fig:bottplot}
\end{figure*}

@@

%The field of exoplanet research has made great advances in the last few decades. 
%The characterisation of Earth-like planets in the habitable zones of their stars is now at the frontier of what we can achieve. 
The Habitable Worlds Observatory (\HWO) will perform a dedicated direct imaging survey of nearby stars, searching for rocky planets in the habitable zone. 
This space-based coronagraphic telescope will detect and characterise exoplanets using their reflection spectrum. 
Complementary information on the habitability of these worlds can be obtained by observing the change in the planet's brightness at different planetary phases.

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, the presence of liquid surface water can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star, for which a rocky planet may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the incident stellar flux that influence a planet's surface temperature and ability to hold liquid surface water, such as the atmospheric surface pressure and geological activity. 
%
Establishing the presence of liquid surface water on an exoplanet would therefore provide important context for the interpretation of possible biosignatures.

Variations of the flux and degree of polarization as an exoplanet orbits its host star---phase variations---could reveal the most distinct signatures of habitability: 
specular reflection (a.k.a. glint) from water oceans, or optical phenomena from scattering by condensates---like the rainbow.
%
However, depending on the architecture of a planetary system, some of these features can be obscured by the inner working angle (\IWA) of a telescope's coronagraph.
%
%Providing additional context to the habitability of an exoplanet would be the presence of a water cycle, as evidenced by the identification of a  as detailed in the next sections.

@@

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off a smooth liquid surface. 
While glint is a familiar phenomenon on Earth, there is only one other body in the Solar System where it has been observed: the Cassini orbiter captured the glint of sunlight reflecting off the liquid hydrocarbon lakes near Titan's North Pole as it was observing the moon at crescent phases \citep{2010GeoRL..37.7104S}.
In the total flux, the glint signal is indeed expected to be most prominent at large planetary phase angles.
As shown in Figure \ref{fig:bottplot}, the maximum degree of polarization of the glint occurs at the Brewster angle, which is determined by the refractive indices at the interface; between air and liquid water this occurs at a scattering angle of \qty{53}{\degree} \citep[i.e., at a planetary phase angle of \qty{106}{\degree}; see, e.g.,][]{2008Icar..195..927W,Zugger_2010,treesandstam2019}.
This angle therefore varies slightly with the wavelength and the composition of the ocean and the atmosphere.
\looseness=-1

In general, the glint on a planet will be strongest at red wavelengths because atmospheric Rayleigh scattering and attenuation are smallest \citep{Zugger_2011}. The stronger glint signature at longer wavelengths has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}. %A degeneracy between polar brightness and glint is also smaller at redder wavelengths \citep{2012ApJ...752L...3C}.
%
The width across phase angles of the glint feature in total and polarized flux generally broadens with increasing wave height \citep{2008Icar..195..927W, kopparla2018, treesandstam2019, trees2022}, which depends strongly on the wind speed over the surface, as described by \citet{CoxMunk1954}.
%

Glint provides evidence of specular reflecting surface bodies, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for the presence of liquid water.
%
At large phase angles, the forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}, which motivates probing the rainbow angle for context, as described below.
%
This confusion can be resolved by measuring polarized phase curves at multiple wavelengths %in polarized light (the polarized colour changes from blue, through white, to red), which only occurs when the glint contributes to the signal and at a planetary scattering angle depending on the cloud coverage fraction 
\citep{treesandstam2019}, as shown in \cref{fig:bottplot}.
%The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
%angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
%While the glint feature can only be probed when a planet is at crescent (i.e., large phase angles), \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry, as the glint changes the difference in polarization between continuum wavelengths and absorption bands. 
%This could be applied when the planet is near quadrature, and thus more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Rainbows, cloudbows and glories}

Rainbows, cloudbows, glories and other optical phenomena from light that is scattered by atmospheric condensates can dramatically influence the shape of a planet's total flux and polarized phase curves and provide a wealth of information about exoplanets \citep{karalidi2012rainbow, stam2008,Bailey2007,2014A&A...566L...1G}.%
\footnote{
    Rainbows occur when light is scattered by rain droplets that are large compared to the wavelength of the light. 
    Cloudbows refer to features occurring at the same phase angles from smaller cloud particles, with the spread in colours invisible to the human eye making them appear white \citep[see][for further discussion]{Bailey2007}.  
}
%
In our own Solar System, the polarized back scatter from Titan provided evidence that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, the polarized phase curve of Venus at three wavelengths showed not only that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture but also constrained the droplet size distribution \citep{hansenhovenier1974}.
%
The glory feature in Venus's total flux phase curve was later used to rule out potential solutions for the unknown UV absorber \citep{petrova2018glory}.

%In unpolarized light, the scattering by cloud particles can dramatically change the shape of a planet's phase curve compared to that of a Lambertian reflecting planet.
%
While the glory and rainbow are technically visible both in the total flux
and the polarized flux phase curves in the presence of spherical cloud and/or haze particles, their strongly polarized nature coupled with the suppression of signals from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized flux than in total flux \citep{karalidi2011, stam2008, treesandstam2019}.
%
The rainbow angle depends on the refractive index of the scattering media, and thus depends on their composition and the wavelength. 
%
The wavelength dependence itself also depends on the particles' size distribution, and the size distribution affects the angular shape of the rainbow \citep{karalidi2011}.
%
Because the rainbow feature is narrower than the glint due to the surface reflection, it would allow for a more exact determination of the composition of the condensates.
%
Note that the scattering by clouds that consist of solid particles, such as ice crystals, will produce other optical features, such as halos, particularly at small phase angles \citep{2022JQSRT.29008313M,Bailey2007,hansentravis1974, karalidi2012rainbow}, that also depend on the composition and shape of the particles \citep[see, e.g.,][for simulations of halos due to carbon dioxide ice crystals]{2023GeoRL..5003457D}.

In this paper, we define our optimal angle for the detection of a rainbow as that of the peak of the water droplet rainbow at visible wavelengths, which is around a phase angle $\alpha$ of \qty{40}{\degree} (see \cref{fig:bottplot}).
%
Pushing observations to angles closer to the star (smaller phase angles, i.e., larger scattering angles) would allow the determination of the full shape of this rainbow, and enable the identification of rainbows due to the scattering of light by other possible HZ condensates.
%
The glory can allow further characterization of condensates as its angle and shape also depend on the complex refractive index, but this feature occurs within a phase angle of \qty{20}{\degree}, thus very close to full phase when the on sky separation is small \citep{hansentravis1974}, as can also be seen in \cref{fig:bottplot}.

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep{decadal} recommended the first in a new \enquote{Great Observatories} program, a telescope with the capability to detect signatures of habitability on about 25 HZ planets.
This requires an instrument with a coronagraph capable of high contrast imaging at optical to near infrared wavelengths.
Following the release of this survey, NASA recently announced the start of the development of the \HWO.

The precursor technology recommended by the survey lists \enquote{direct imaging to probe polarized ocean glint on terrestrial planets} as one of the priority capabilities \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
The performance and precise characteristics of \HWO (on-axis or off-axis, diameter, type of segmentation, type of coronagraph) are still to be determined.
%
The development will be heavily informed by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
The Inner Working Angle (\IWA) of the resulting telescope will depend on these characteristics, which in turn will have a significant impact on the expected exo-Earth yield \citep{Stark2019_exoplanetyield}.

The Extreme Coronagraph for Living Planetary Systems (ECLIPS) coronagraph for LUVOIR was a proposed \qtyrange{200}{2000}{\nano\meter} instrument for exoplanet characterization, with a similar coronagraph, albeit operating between \num{450} and \qty{1800}{\nano\meter}, planned for HabEx.
%
POLLUX, a proposed instrument for LUVOIR would detect polarized light from hot Jupiters at \qty{300}{\nano\meter} \citep{Bouret2018_pollux}.
%
However, because of the challenges associated with the design of a coronagraph in the UV, it appears to be unlikely that \HWO will include a UV high-contrast instrument sensitive to Earth-like planets.

@@

In this paper, we quantify the planetary phase angles that would be accessible for direct observations of hypothetical terrestrial planets orbiting in the habitable zones of \HWO target stars.
%
In \cref{subsec:2.1}, we describe the target list, in \cref{subsec:2.2}, we derive expressions for the range of phase angles given a few different limits, in \cref{subsec:2.3}, we describe Monte Carlo simulations to marginalize over the unknown orbital inclination and eccentricity of planets in these systems, and in \cref{subsec:2.4}, we outline how the planet/star contrast changes along a planet's orbits.  
%
We present and discuss our results in \cref{sec:3}, and conclude in \cref{sec:4}. 
It is important to note that the phases angles selected for the optical phenomena throughout this paper are approximate, that phenomena overlap, and that their maximum peak will depend on the chemical species and wavelength.

\subsection{Angle conventions}
The conventions for angle names vary throughout the literature on exoplanet phase curves. 
In this paper, we define the \emph{scattering angle} as the local divergence in the propagation direction of the light (large angles for back-scattering, small angles for forward-scattering); the \emph{phase angle} $\varphi$ as the viewer-star-planet angle, which is zero at full phase and 180 at new phase, and which is the complement of the scattering angle. 
A planet with a face-on orbit will thus have a phase angle of \qty{90}{\degree} everywhere along its orbit. 
In some plots, we \enquote{fold} the half orbit so that we measure \emph{degrees from quadrature} (so that both new and full phase are at maxima of \qty{90}{\degree} for edge-on orbits), which is helpful when considering the observability of the planet given a certain inner working angle. 
We also note that we are considering here the (linearly) polarized flux, which considers the intensity of the polarization and reflectance (illumination) of the planet. 
Additional information could be garnered from full Stokes characterization.

An exoplanet in an edge-on orbit moves through phase angles ranging from \qty{0}{\degree} (full phase), through \qty{90}{\degree} (quadrature), and to \qty{180}{\degree} (the planet's nightside is towards the observer) (cf.\ \cref{fig:bottplot}), while planets in higher inclination orbits will pass through a subset of these phase angles.

@@

Most planets are not in edge-on orbits as seen from Earth.
For an orbit with an inclination angle $i$, we can identify the following two regimes:% (see also \cref{fig:orb-grid}):
\begin{enumerate}
    \item The coronagraph does not obscure any part of the sky-projected orbit: 
    the maximum phase angle coverage depends solely on $i$. 
        This holds for orbits that are relatively face-on. 
    \item The coronagraph does obscure parts of the sky-projected orbit: the phase angle coverage depends only on the \IWA and the orbital semi-major axis $a$. 
\end{enumerate}

@@

Using \cref{eq:scattering_angle}, we deduce the following for inclined circular orbits: 
\begin{equation}
\label{eq:Delta_phi_max}
    \Delta \varphi = 
    \begin{cases}
        2 i & \textrm{for} \cos(i) > \frac{\mathrm{IWA}\; d_* }{a} \\ 
        2 \cos^{-1}\left(\dfrac{\mathrm{IWA}\cdot d_* }{a}\right) & \textrm{for} \cos(i) < \frac{\mathrm{IWA}\; d_* }{a}
    \end{cases} \,.
\end{equation}

\subsection{Monte Carlo simulations}
\label{subsec:2.3}

>>>>

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\HWO), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight reflected by an exoplanet can reveal ocean glint, rainbows and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, and \HWO' coronagraph may obscure the phase angles where such optical features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
For planets in near edge-on orbits, accessible phase angles are limited by coronagraphic obscuration. 
%
We use \HWO'~catalogue of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering. 
%
We find that any exo-Earths in planetary systems in this catalogue will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The rainbow due to water clouds at scattering angles of \qtyrange{140}{160}{\degree} would be accessible to \HWO\ in $\sim\num{42}$ systems, while ocean glint signature at scattering angles of $\qtyrange{50}{70}{\degree}$ would be accessible in $\sim\num{14}$ systems, assuming an \IWA of \qty{62}{\mas} ($3\,\lambda/D$ for a $D=\qty{6}{\meter}$ telescope at $\lambda=\qty{600}{\nano\meter}$).
%
Improving the \IWA to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases slightly for eccentric orbits.  %and scales approximately inversely with the \IWA.  
\end{abstract}

@@

@@

%The field of exoplanet research has made great advances in the last few decades. 
%The characterisation of Earth-like planets in the habitable zones of their stars is now at the frontier of what we can achieve. 
%The Habitable Worlds Observatory (\HWO) will perform a dedicated direct imaging survey of nearby stars, searching for rocky planets in the habitable zone. 
%This space-based coronagraphic telescope will detect and characterise exoplanets using their reflection spectrum. 
%Complementary information on the habitability of these worlds can be obtained by observing the change in the planet's brightness at different planetary phases.

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, its presence can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star, for which a rocky planet may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the incident stellar flux that influence a planet's ability to hold liquid surface water, such as the atmospheric pressure and geological activity. 
%
Detecting liquid surface water on an exoplanet would thus provide important context for the interpretation of possible biosignatures.
%
Variations of the flux and degree of polarization as an exoplanet orbits its host star---phase variations---could reveal scattering phenomena from condensed water \citep[e.g., rainbows;][]{bailey2007} or specular reflection, i.e. glint, from water oceans \citep[][]{2008Icar..195..927W}.
%
However, depending on a planetary system's architecture, such features can be obscured by the inner working angle (\IWA) of a telescope's coronagraph, shown schematically in Fig.~\ref{fig:annotated-orbit}.
%
%Providing additional context to the habitability of an exoplanet would be the presence of a water cycle, as evidenced by the identification of a  as detailed in the next sections.

@@

\subsection{Angle conventions}
%The conventions for angle names vary throughout the literature on exoplanet phase curves. 
As shown in Fig.~\ref{fig:annotated-orbit}, we define the \emph{scattering angle} $\varphi$ as the planet--star--observer angle 
%local divergence in the propagation direction of the light 
(large angles for back-scattering, small angles for forward-scattering) and the supplementary \emph{phase angle} $\alpha$ (zero at full phase and 180$^\circ$ at new phase). 
A planet on a perfectly face-on orbit is always seen at quadrature, $\varphi = \alpha = 90^\circ$, while a planet on an inclined orbit will exhibit a wider range of viewing angles; only perfectly edge-on orbits pass through all possible scattering angles. 
%a phase angle of \qty{90}{\degree} everywhere along its orbit. 
In some plots, we report \emph{degrees from quadrature}, so that both new and full phase are at \qty{90}{\degree} for edge-on orbits.
%, which is helpful when considering the observability of the planet given a certain inner working angle.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includestandalone{tikz/annotated-orbit}
    \caption{
        Illustration of a planet in an edge-on, circular orbit showing how a coronagraph can obscure the exoplanet at small and large phase angles, and thus prevent observations of characteristic scattering features in an exoplanet's
        phase curve. 
        The scattering angle, $\varphi$, is shown in red and the supplementary phase angle, $\alpha$, is shown in \todo{blah}. Also see Figure~\ref{fig:scattering-angle}
    }
    \label{fig:annotated-orbit}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is the specular (Fresnel) reflection of light off a smooth liquid surface. 
While glint is a familiar phenomenon on Earth, there is only one other body in the Solar System where it has been observed: the Cassini orbiter captured the glint of sunlight reflecting off the liquid hydrocarbon lakes near Titan's North Pole as it was observing %the moon 
Titan at crescent phases \citep{2010GeoRL..37.7104S}.
In the total flux, the glint signal is indeed expected to be most prominent at large planetary phase angles.
%As shown in Figure \ref{fig:bottplot}, 
The maximum degree of polarization of the glint occurs at the Brewster angle, which is determined by the refractive indices at the interface; between air and liquid water this occurs at a scattering angle of \qty{53}{\degree} \citep[i.e., at a planetary phase angle of \qty{106}{\degree}; see, e.g.,][]
%{2008Icar..195..927W,Zugger_2010,treesandstam2019}
{Zugger_2010,treesandstam2019}.
This angle therefore varies slightly with the wavelength and the composition of the ocean and the atmosphere.
The angular width of the glint in total and polarized flux generally broadens with increasing wave height \citep{2008Icar..195..927W, kopparla2018, treesandstam2019, trees2022}.
\looseness=-1

In general, the glint on a planet will be strongest at red wavelengths because there atmospheric Rayleigh scattering and attenuation are smallest \citep{Zugger_2011}. This has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}. %A degeneracy between polar brightness and glint is also smaller at redder wavelengths \citep{2012ApJ...752L...3C}.
%
%The width across phase angles of the glint feature in total and polarized flux generally broadens with increasing wave height \citep{2008Icar..195..927W, kopparla2018, treesandstam2019, trees2022}, which depends on the surface wind speed, as described by \citet{CoxMunk1954}.
%

Glint provides evidence of specular reflecting surfaces, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for liquid water.
%
At large phase angles, the forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}.
%
This confusion can be resolved by measuring polarized phase curves at multiple wavelengths
\citep{treesandstam2019}, as shown in \cref{fig:bottplot}, or by probing 
the rainbow, see below. %in polarized light (the polarized colour changes from blue, through white, to red), which only occurs when the glint contributes to the signal and at a planetary scattering angle depending on the cloud coverage fraction 
%\citep{treesandstam2019}, as shown in \cref{fig:bottplot}.
%The cloud coverage fraction on ocean worlds could be constrained with observations spanning the ultraviolet to the near-infrared by finding the 
%angle at which the planet's polarization phase curves at various wavelengths intersect \citep{treesandstam2019}.
%
%While the glint feature can only be probed when a planet is at crescent (i.e., large phase angles), \citet{trees2022} showed that reflection by a liquid surface body can also be detected with spectropolarimetry, as the glint changes the difference in polarization between continuum wavelengths and absorption bands. 
%This could be applied when the planet is near quadrature, and thus more easily probed when using coronagraphs.
%how use to map, 
%seen in Earthshine
%strong signal for exoplanets
% disentangled from cloud coupled with rainbow
%disentangled from lattitude albedo coupled in polarization

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-1-bott-plot.pdf}
    \script{create-figure-1-bott-plot.py}
    \caption{
        The phase curves of the normalized linearly polarized flux of a cloudy planet with a wavy ocean surface and an Earth-like atmosphere in an edge-on orbit; adapted from \citet{treesandstam2019}.  
        %(the planet goes through all possible phase angles throughout its orbit). 
        The wind speed over the ocean, which determines
        the wave height \citep[][]{CoxMunk1954}, is \qty{7}{\meter\per\second}. 
        The clouds consist of spherical water droplets. 
        The cloud coverage fraction is 0.5 and while the  
        cloud pattern is patchy, the glint is always cloud-free. 
        %to maximize the glint signature \citep[see also Fig. 9 of][from which we took the mean signal of 300 planets with randomly generated patchy cloud patterns + 2 standard deviations, representative for cloud-free glint scenarios; the \enquote{noise} at the larger phase angles and longer wavelengths is due to the variation in the location of cloud patches]{treesandstam2019}. 
        %The polarized flux is normalized to the total (unpolarized + polarized) flux (it is thus in fact the degree of polarization). 
        % Note that the reference plane for the polarized flux is chosen such that only the polarized components perpendicular or parallel to the plane differ significantly from zero.
        % and the latter
        The fluxes are normalized such that the total (unpolarized + polarized) flux equals the planet's geometric albedo at phase angle $\alpha = 0^\circ$.
        % The cloudy scenario includes a low-altitude, Earth-like water 
        % cloud deck. 
        Below the plot, we show the scattering angles of glories, rainbows, Rayleigh scattering, and ocean glint. Note that the features in the actual curves partly overlap with each other.
        %Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (phase angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest phase angles).
    }
    \label{fig:bottplot}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Rainbows, cloudbows, and glories}

Rainbows, cloudbows, glories and other scattering phenomena can dramatically influence the shape of a planet's photometric and polarimetric phase curves and provide a wealth of information about exoplanets \citep{karalidi2012rainbow, stam2008,bailey2007,2014A&A...566L...1G}.%
\footnote{
    Rainbows occur when light is scattered by rain droplets that are large compared to the wavelength of the light. 
    Cloudbows refer to features occurring at the same phase angles from smaller cloud particles, with the spread in colours invisible to the human eye making them appear white \citep[for further discussion, see][]{bailey2007}.  
}
%
In the Solar System, the polarized back-scatter from Titan indicated that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, Venus's polarized phase curve at three wavelengths showed that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture and constrained the droplet size distribution \citep{hansenhovenier1974}.
%
The glory feature in Venus's flux phase curve was later used to rule out various species as the unknown UV absorber \citep{petrova2018glory}.
%In unpolarized light, the scattering by cloud particles can dramatically change the shape of a planet's phase curve compared to that of a Lambertian reflecting planet.
%
Glories and rainbows are technically visible both in total and polarized flux phase curves in the presence of spherical cloud and/or haze particles, but their strongly polarized nature coupled with the suppression of other polarized features at gibbous and full phases make them more easily characterized with polarimetry than with photometry \citep{karalidi2011, stam2008, treesandstam2019,trees2022}.
%

% The rainbow angle depends on the refractive index of the scattering media, and thus depends on their composition and the wavelength. 
% %
% The wavelength dependence itself also depends on the particles' size distribution, and the size distribution affects the angular shape of the rainbow \citep{karalidi2011}.
% %
% Because the rainbow feature is narrower than the glint due to the surface reflection, it would allow for a more exact determination of the composition of the condensates.
% %
% Note that the scattering by clouds that consist of solid particles, such as ice crystals, will produce other optical features, such as halos, particularly at small phase angles \citep{2022JQSRT.29008313M,Bailey2007,hansentravis1974, karalidi2012rainbow}, that also depend on the composition and shape of the particles \citep[see, e.g.,][for simulations of halos due to carbon dioxide ice crystals]{2023GeoRL..5003457D}.

In this paper, we define our optimal angle for the detection of a rainbow as that of the peak of the water droplet rainbow at visible wavelengths, which is around a phase angle $\alpha$ of \qty{40}{\degree} (see \cref{fig:bottplot}).
%
Pushing observations to angles closer to the star (smaller phase angles, i.e., larger scattering angles) would allow the determination of the full shape of the rainbow, and enable the identification of rainbows due to the scattering of light by other condensates.
%
The glory can allow further characterization of condensates as its angle and shape also depend on the refractive index, but this glory occurs within a phase angle of \qty{20}{\degree}, when the on-sky separation is small, as can also be seen in \cref{fig:bottplot}. We note that we are considering here only linearly polarized fluxes, not circularly polarized fluxes. 
%which considers the intensity of the polarization and reflectance (illumination) of the planet. 
%Additional information could be garnered from a full Stokes characterization.

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability on about 25 HZ planets.
This requires an instrument with a coronagraph capable of high-contrast imaging and low-resolution spectroscopy at optical to near-infrared wavelengths. 
%It will measure the planet's reflected light at low spectral resolution to identify its atmospheric constituents.
The precursor technology recommended by the survey lists \enquote{direct imaging to probe polarized ocean glint on terrestrial planets} as a priority capability \citep[Box E.1 in][]{decadal} for ground and space-based observatories.

%
Following the release of this report, NASA recently announced the start of the development of the Habitable Worlds Observatory (hereafter ``\HWO'').
%
The performance and characteristics (on- or off-axis, diameter, type of segmentation and coronagraph, \IWA) of \HWO will significantly 
impact the expected exo-Earth yield 
\citep{Stark2019_exoplanetyield}
and are still to be determined.
%
They will likely be heavily informed by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
%\HWO' \IWA and thus the expected exo-Earth yield and characterization capabilities depend on these characteristics \citep{Stark2019_exoplanetyield}.

% The Extreme Coronagraph for Living Planetary Systems (ECLIPS) coronagraph for LUVOIR was a proposed \qtyrange{200}{2000}{\nano\meter} instrument for exoplanet characterization, with a similar coronagraph, albeit operating between \num{450} and \qty{1800}{\nano\meter}, planned for HabEx.
% %
% POLLUX, a proposed instrument for LUVOIR would detect polarized light from hot Jupiters at \qty{300}{\nano\meter} \citep{Bouret2018_pollux}.
% %
% However, because of the challenges associated with the design of a coronagraph in the UV, it appears to be unlikely that \HWO will include a UV high-contrast instrument sensitive to Earth-like planets.

@@

In this paper, we quantify the planetary phase angles that would be accessible for direct observations of hypothetical terrestrial planets orbiting in the habitable zones of \HWO target stars.
%
In \cref{subsec:2.1}, we describe the target list, in \cref{subsec:2.2}, we derive expressions for the range of phase angles given a few different limits, in \cref{subsec:2.3}, we describe Monte Carlo simulations to marginalize over the unknown orbital inclination and eccentricity of planets around these stars, and in \cref{subsec:2.4}, we outline how the photometric and polarimetric planet/star contrast changes along a planet's orbit.  
%
We present and discuss our results in \cref{sec:3}, and conclude in \cref{sec:4}.

%It is important to note that the phases angles selected for the optical phenomena throughout this paper are approximate, that phenomena overlap, and that their maximum peak will depend on the chemical species and wavelength.

%An exoplanet in an edge-on orbit moves through phase angles ranging from \qty{0}{\degree} (full phase) through \qty{90}{\degree} (quadrature) to \qty{180}{\degree} (the planet's nightside is towards the observer; cf.\ \cref{fig:bottplot}), while planets in more inclined orbits will pass through a subset of these phase angles.

@@

Most planets are not in edge-on orbits as seen from Earth.
For an orbit with an inclination angle $i$ ($i=0^\circ$ is face-on), we can identify the following two regimes:% (see also \cref{fig:orb-grid}):
\begin{enumerate}
    \item The coronagraph does not obscure any part of the sky-projected orbit: 
    the maximum phase angle coverage depends solely on $i$. 
        This holds for orbits that are relatively face-on (small $i$). 
    \item The coronagraph does obscure parts of the sky-projected orbit: the phase angle coverage depends only on the \IWA and the orbital semi-major axis $a$, following Eq.~\ref{eq:scattering_angle}. 
\end{enumerate}

@@

%Using \cref{eq:scattering_angle}, we deduce the following 
The range of accessible scattering angles is then 
\begin{equation}
\label{eq:Delta_phi_max}
    \Delta \varphi = 
    \begin{cases}
        2 i & \textrm{for} \cos(i) > \frac{\mathrm{IWA}\; d_* }{a} \\ 
        2 \cos^{-1}\left(\dfrac{\mathrm{IWA}\cdot d_* }{a}\right) & \textrm{for} \cos(i) < \frac{\mathrm{IWA}\; d_* }{a}
    \end{cases} \,
\end{equation}

\subsection{Monte Carlo simulations of accessible orbits}
\label{subsec:2.3}
